Contracts for Contracts

### 2023-06-27 paper/formal_model.tex
more syntax

The body of functions is simply a statement.
Statements can be a type-and-identifier declaration (without providing an initial
value), an expression (e.g. calling a function for its side effects), a
composition of two statements, an assignment, a \Keywd{return} statement, a \Keywd{revert} statement,
or a conditional statement.
The left-hand side part of assignments (i.e. assignable) can be either
a type-and-type declaration, an identifier, or a projection of mappings or structs.
Since \corelang permits both mappings and structs, which can recursively referring
other types, projection allows use to directly access nested data structures,
e.g. \code{v.field1[0].field2}.
A \Keywd{revert} statement aborts the execution, reverting any changes made to
the blockchain state.

Expressions could be an identifier, a literal value (e.g. numbers), a binary
operation, projections of mappings or structs, address/function calls.
Address and function calls have form $\AddrCall{e_1}{(x: e_2)^*}{e_3}$,
where $e_1$ is the target callee expression,
$(x: e_2)^*$ are call options (e.g. specifying the gas budget),
and $e_3^*$ are call arguments.
An ordinary function call or struct constructor call omits the call option
part, consisting only of the function name and argument expressions.

Specifications attached to function definitions first introduce bindings for
the guarded function, including the function name, option parameters,
ordinary parameters, and returned values.
The precondition and postcondition are denoted in the \code{requires}-clause and
\code{ensures}-clause, respectively.
In practice not all functions need to be specified with conditions.
In such cases, a no-op specification is expressed where the precondition and
postcondition are simply a boolean expression \code{true}.
In the \code{where}-clause, programmers can specify the condition for addresses
appeared as arguments or returned values for the parent call, recursively using
the same syntax.
When the specification is attached to an address, we allow very flexible ways
to specify the callee. For example, if a struct argument contains an address
value that we would like to guard:
\begin{lstlisting}
// x is a struct argument whose field y is an address
... where { x.y.f requires $\mathit{precond}$ ensures $\mathit{postcond}$ }
\end{lstlisting}
we can use the projection syntax to specify
the condition when address \code{x.y} is called on function \code{f}.

\subsection{Static Semantics}

@@

\begin{figure}[t]
  \judgement{Function}{$\Gamma \vdash \sigma: t$}
  %\begin{alignat*}{2}
    %\mathit{param}(\SpecCall{f}{(x : y)^*}{x^*}{y^*} ...) & = 1
  %\end{alignat*}

  \infrule[f-spec]{
    \mathit{target}(\sigma) = f \\
    \Gamma' = \Gamma \cup
              \{ \mathit{param}(\sigma)_i : \mathit{type}(d_i) \}^{i \in 1..m} \\
    \Gamma'' = \Gamma' \cup
              \{ \mathit{ret}(\sigma)_j : t_j \}^{j \in 1..n} \\
    \Gamma' \vdash \mathit{pre}(\sigma) : \Keywd{bool} \andalso
    \Gamma'' \vdash \mathit{post}(\sigma) : \Keywd{bool} \\
    \Gamma \vdash \mathit{addr}(\sigma)_i : \note{type-of-call}
  }{
    \Gamma \vdash \sigma : \FunType{f}{d_i^{i \in 1..m}}{m}{t_j^{j \in 1..n}}
  }

  \infrule[c-top]{
    \forall i \in 1..n,\ \Gamma_0 \cup \Gamma \vdash \F_i \\
    \Gamma = \{ \mathit{id}(d_j) : \mathit{type}(d_j) \}^{j \in 1..m} \cup
             \{ \mathit{id}(\F_i) : \mathit{type}(\F_i) \}^{i \in 1..n}
  }{
    \Gamma_0 \vdash \Contract{c}{d_i^{i \in 1..m}}{{\F}_i^{i \in 1..n}}
  }

  % \Gamma_0 initial context, containing typings for built-in terms

  \caption{Static semantics (excerpt) of \corelang specifications. Only checkings relevant to specifications are shown.}
  \label{fig:static}
\end{figure}

>>>>

We can also define interfaces at top-level, which only contain function
declarations. Interfaces are useful when invoking functions on addresses,
providing expected types of arguments and returned values of the address call.

The body of functions is simply a statement.
Statements can be a type-and-identifier declaration (without providing an initial
value), an expression (e.g. calling a function for its side effects), a
composition of two statements, an assignment, a \Keywd{return} statement, a \Keywd{revert} statement,
or a conditional statement.
The left-hand side part of assignments (i.e. assignable) can be either
a type-and-type declaration, an identifier, or a projection of mappings or structs.
Since \corelang permits both mappings and structs, which can recursively refer to
other types, projection allows use to directly access nested data structures,
e.g. \code{v.field1[0].field2}.
A \Keywd{revert} statement aborts the execution, reverting any changes made to
the blockchain state.

Expressions could be an identifier, a literal value (e.g. numbers), a binary
operation, projections of mappings or structs, address/function calls.
Address and function calls have form $\AddrCall{e_1}{(x: e_2)^*}{e_3}$,
where $e_1$ is the target callee expression,
$(x: e_2)^*$ are call options (e.g. specifying the gas budget),
and $e_3^*$ are call arguments.
An ordinary function call, struct constructor call, or casting omits the call
option part, consisting only of the function name and argument expressions.

Specifications attached to function definitions first introduce bindings for
the guarded function, including the names for the guarded callee, call options,
ordinary arguments, and returned values. The following snippet shows an example
\[
  \overbrace{\mathit{fp^*}}^{\text{callee}}\Keywd{\{}\underbrace{(x : y)^*}_{\text{call options}}\Keywd{\}(}\overbrace{x^*}^{\text{arguments}}\Keywd{)} ~\Keywd{returns (}\underbrace{y^*}_{\text{returned values}}\Keywd{)}
\]
The precondition and postcondition are denoted in the \code{requires}-clause and
\code{ensures}-clause, respectively.
In practice not all functions need to be specified with conditions.
In such cases, a no-op specification is expressed where the precondition and
postcondition are simply a boolean expression \code{true}.
In the \code{where}-clause, programmers can specify the condition for addresses
appeared as arguments or returned values for the parent call, recursively using
the same syntax.
When the specification is attached to an address, we allow very flexible ways
to specify the callee. For example, if a struct argument contains an address
value that we would like to guard:
\begin{lstlisting}
// x is a struct argument whose field y is an address
... where { x.y.f requires $\mathit{precond}$ ensures $\mathit{postcond}$ }
\end{lstlisting}
we can use the projection syntax to specify
the condition when address \code{x.y} is called on function \code{f}.

\subsection{Static Semantics}
\input{fig:static}

@@
